\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-08-01 docs/parco13/paper.tex
Crop a pdf
Tweak the paper for resubmission

\begin{abstract}
We present a new hybrid paradigm for parallel adaptive mesh refinement (AMR)
that combines the scalability and lightweight architecture of tree-based AMR
with the computational efficiency of patch-based solvers for hyperbolic
conservation laws.  The key idea is to interpret each leaf of the AMR hierarchy
as one uniform compute patch in $\sR^d$ with $m^d$ degrees of freedom, where $m$ is
customarily between 8 and 32.  Thus, computation on each patch can be optimized
for speed, while we inherit the flexibility of adaptive meshes.  In our work we
choose to integrate with the \pforest AMR library since it allows us to compose
the mesh from multiple mapped octrees and enables the cubed sphere and other
nontrivial multiblock geometries.  We describe aspects of the parallel
implementation and close with scalings for both MPI-only and OpenMP/MPI hybrid
runs, where the largest MPI-only run uses 16,384 CPU cores.
\end{abstract}

@@

\begin{figure}
\begin{center}
%  \includegraphics[angle=-90,width=.8\columnwidth]{../figs_misc/omp}
%  \\[1ex]
  \includegraphics[angle=-90,width=.75\columnwidth]{../figs_misc/advance_strong_omp}
\end{center}
\caption{Strong scaling of MPI/OpenMP parallelism for advection on a
  uniformly refined mesh with 16 threads per node.
  %Top: run time of advection Riemann problem kernel in both an optimized
  %inlined interface and a generic function pointer interface via \manyclaw.
  %Bottom:
  %MPI + OpenMp hybrid parallelization of the time integration.
}
\label{fig:ompscale}
\end{figure}%

@@

We have examined weak scalability as well by using four times as many processes
for each level increase, from level 6 at 64 cores to level 11 at 16,384.  Up to
4,096 cores, the exchange times stay consistently below 16\% of the time
integration and the time in seconds per time integration step varies between
.0578 and .0602, which yields a parallel efficiency of 96\%.  For the
data point at 16,384 cores, the efficiency is still 82\%.

@@

\begin{figure}
  \begin{center}
    \hspace{.1\columnwidth}
    \begin{minipage}[c]{.35\columnwidth}
      \includegraphics[width=.99\columnwidth]{../figs_misc/sphere_partition5}
    \end{minipage}
    \hspace{.05\columnwidth}
    \begin{minipage}[c]{.4\columnwidth}
    \begin{tabular}{r|r}
       $P$ & time int./step \\
      \hline
       16 & 1.115 \\
       32 & 0.557 \\
       64 & 0.297 \\
      256 & 0.092 \\
    \end{tabular}
  \end{minipage}
\end{center}
\caption{Left: Partition of the spherical Example 2 between five MPI processors, indicated
  by the color scale, at simulated time $T = 1$.  The partition shows the mesh equator,
  while the refinement is invariant under the number of processors and the partition
  and follows the front, which is not aligned with the equator.  Right: Strong scaling
  for a uniform run at level 8.}
\label{fig:spherical}
\end{figure}
To showcase a multiblock connectivity, we include results for the spherical Example 2 in
\Fig{spherical}, together with a strong scaling table from 16 to 256 MPI ranks.

@@

We have presented a novel hybrid paradigm for AMR that achieves good
scaling both in the distributed and shared memory aspects of current
supercomputers.
%
% By
% abstracting a leaf in the AMR hierarchy as the atomic unit of computation, we
% are able to use almost arbitrary low level solvers.
%
Our implementation integrates an MPI-based forest-of-octrees adaptive
meshing code, \pforest, with a numerical solver for hyperbolic
conservation laws, \clawpack/\manyclaw,
that implements threaded parallelism for a single compute patch.
%All topological operations on the mesh are handled by \pforest,
%while all numerical operations are handled by the \clawpack instance.
This combination naturally enables MPI/threaded hybridization,
where our biggest MPI run scales to 16,384 CPU cores.

%
% We demonstrate scalability of this approach in various ways, where the MPI-only variant
% scales weakly up to 16,384 codes, and strong scalability is linear for up to 256 cores
% in the MPI-only and 128 in the threaded implementation, and our results that higher
% core counts are easily possible.
%
% d to manage the exchanges between them efficiently.  The
% flexibility and scalability of this approach provides an ideal workspace to
% explore further algorithms and physics problems requiring new solvers at high
% resolutions.
%
%the versatility of this approach the modular development of
%scalable codes.

@@

We would like to thank the Texas Advanced Computing Center (TACC) for access to
the Stampede supercomputer under allocations TG-DPP130002 and TG-ASC130001
granted by the NSF XSEDE Program.
The authors acknowledge valuable discussion with Randy LeVeque, Marsha
Berger, and Hans-Petter Langtangen.  We also acknowledge David
Ketcheson and the KAUST sponsored HPC$^3$ numerics workshop at which
the initial phases of this project were first discussed.  The second
author would like to also acknowledge the Isaac Newton Institute
(Cambridge, UK), where much of the preliminary development work for
\forestclaw was done.
The fourth author recognizes Simula National Lab for funding.
The leaf/patch paradigm was independently
presented by B.\ as part of a talk at the SCI Institute, Salt Lake City, Utah,
USA in July 2011.

>>>>

\begin{abstract}
We present a new hybrid paradigm for parallel adaptive mesh refinement (AMR)
that combines the scalability and lightweight architecture of tree-based AMR
with the computational efficiency of patch-based solvers for hyperbolic
conservation laws.  The key idea is to interpret each leaf of the AMR hierarchy
as one uniform compute patch in $\sR^d$ with $m^d$ degrees of freedom, where $m$ is
customarily between 8 and 32.  Thus, computation on each patch can be optimized
for speed, while we inherit the flexibility of adaptive meshes.  In our work we
choose to integrate with the \pforest AMR library since it allows us to compose
the mesh from multiple mapped octrees and enables the cubed sphere and other
nontrivial multiblock geometries.  We describe aspects of the parallel
implementation and close with scalings for both MPI-only and OpenMP/MPI hybrid
runs, where the largest MPI run executes on 16,384 CPU cores.
\end{abstract}

@@

\begin{figure}
\begin{center}
%  \includegraphics[angle=-90,width=.8\columnwidth]{../figs_misc/omp}
%  \\[1ex]
  \includegraphics[angle=-90,width=.75\columnwidth]{../figs_misc/advance_strong_omp-crop}
\end{center}
\caption{Strong scaling of MPI/OpenMP parallelism for advection on a
  uniformly refined mesh with 16 threads per node.
  %Top: run time of advection Riemann problem kernel in both an optimized
  %inlined interface and a generic function pointer interface via \manyclaw.
  %Bottom:
  %MPI + OpenMp hybrid parallelization of the time integration.
}
\label{fig:ompscale}
\end{figure}%

@@

We have examined weak scalability as well by using four times as many MPI
processes
for each level increase, from level 6 at 64 cores to level 11 at 16,384.  Up to
4,096 cores, the exchange times stay consistently below 16\% of the time
integration and the time in seconds per time integration step varies between
.0578 and .0602, which yields a parallel efficiency of 96\%.  For the
data point at 16,384 cores, the efficiency is still 82\%.

@@

\begin{figure}
  \begin{center}
    \hspace{.1\columnwidth}
    \begin{minipage}[c]{.30\columnwidth}
      \includegraphics[width=\columnwidth]{../figs_misc/sphere_partition5}
    \end{minipage}
    \hspace{.05\columnwidth}
    \begin{minipage}[c]{.4\columnwidth}
    \begin{tabular}{r|r}
       $P$ & time int./step \\
      \hline
       16 & 1.115 \\
       32 & 0.557 \\
       64 & 0.297 \\
      256 & 0.092 \\
    \end{tabular}
  \end{minipage}
\end{center}
\caption{Left: Partition of the spherical Example 2 between five
  MPI processes, indicated by the color scale, at simulated time $T = 1$.
  % The  partition shows the mesh equator,
  The refinement is invariant under the parallel partition
  and follows the front, which is not aligned with the equator.  Right: Strong scaling
  for a uniform run at level 8.}
\label{fig:spherical}
\end{figure}
To showcase a multiblock connectivity, we include results for the spherical Example 2 in
\Fig{spherical}, together with a strong scaling table from 16 to 256 MPI ranks.

@@

We have presented the integration of an MPI-based forest-of-octrees adaptive
meshing code, \pforest, with a numerical solver for hyperbolic conservation
laws, \clawpack/\manyclaw, that implements threaded parallelism for a single
compute patch.
%All mesh operations are handled by \pforest,
%while numerics are handled by the \clawpack implementation.
We abstract an interface to the parallel meshing code in order to derive the
schedule of neighbor exchanges, averaging/interpolation, and time integration
on the local patches.
This approach naturally lends itself to MPI/thread hybridization.
Apart from its parallel scalability, we favor the presented strategy
for its modularity, encapsulation, and versality.  Future work
will continue to optimize the parallel neighbor exchange patterns
and investigate a generic handling of arbitrary multi-block geometries.

% We have presented a novel hybrid paradigm for AMR that achieves good
% scaling both in the distributed and shared memory aspects of current
% supercomputers.
% %
% % By
% % abstracting a leaf in the AMR hierarchy as the atomic unit of computation, we
% % are able to use almost arbitrary low level solvers.
% %
% Our implementation integrates an MPI-based forest-of-octrees adaptive
% meshing code, \pforest, with a numerical solver for hyperbolic
% conservation laws, \clawpack/\manyclaw,
% that implements threaded parallelism for a single compute patch.
% %All topological operations on the mesh are handled by \pforest,
% %while all numerical operations are handled by the \clawpack instance.
% This combination naturally enables MPI/threaded hybridization,
% where our biggest MPI run scales to 16,384 CPU cores.
%
% We demonstrate scalability of this approach in various ways, where the MPI-only variant
% scales weakly up to 16,384 codes, and strong scalability is linear for up to 256 cores
% in the MPI-only and 128 in the threaded implementation, and our results that higher
% core counts are easily possible.
%
% d to manage the exchanges between them efficiently.  The
% flexibility and scalability of this approach provides an ideal workspace to
% explore further algorithms and physics problems requiring new solvers at high
% resolutions.

@@

We would like to thank the Texas Advanced Computing Center (TACC) for access to
the Stampede supercomputer under allocations TG-DPP130002 and TG-ASC130001
granted by the NSF XSEDE program.
The authors acknowledge valuable discussion with Randy LeVeque, Marsha
Berger, and Hans-Petter Langtangen.  We also acknowledge David
Ketcheson and the KAUST sponsored HPC$^3$ numerics workshop at which
the initial phases of this project were first discussed.  The second
author would like to also acknowledge the Isaac Newton Institute
(Cambridge, UK), where much of the preliminary development work for
\forestclaw was done.
The fourth author recognizes Simula Research Lab, Norway, for funding.
The leaf/patch paradigm was independently
presented by B.\ as part of a talk at the SCI Institute, Salt Lake City, Utah,
USA in July 2011.
